A finite-state morphological transducer for Kyrgyz

### 2012-03-14 paper/paper.tex
blzizz

One of the morphotactic challenges met in defining a finite-state transducer for Kyrgyz is that many finite verb forms have ``irregular'' negative forms.  While the paradigms are completely regular, the negative morphotactics are not regularly derived from the affirmative forms.  There are also several different ``regular'' patterns of alternation.  Listed in table~\ref{irregnegs} are a couple examples of a few finite verb paradigms with their negative forms.

@@

However, instead of doing complicated splitting of the case continuation lexica following the third person possessive suffix to sometimes insert /н/, we decided to proceed under the premise that /н/ was instead underlying in this suffix (and a couple others!) and got deleted in the nominative, accusative, and genitive.  To accomplish this, a \{n\} archiphoneme was added to the 3rd person possessive suffix, creating an underlying form of \{S\}\{I\}\{n\}.  Phonological rules were implemented in twolc which deleted \{n\} when followed by nothing (for nominative), another set of rules that deleted the accusative \{N\}\{I\} after \{n\} and a morpheme boundary, and a rule that deleted \{n\} when followed by a morpheme boundary and the genitive \{N\}\{I\}н.\footnote{Because a general possessive suffix that can follow personal possessive suffixes can also behave like the genitive in this respect, the rule is actually more general.}

@@

\subsection{Desonorisation}
Desonorisation \citep{washington10} happens to /N/ when it follows any consonant, and to /L/ when it follows a consonant of equal or lower sonority (i.e., /л/, nasals (/м/, /н/, /ң/), and obstruents, but not /й/, /р/, or vowels).  The resulting surface consonant is [д] for both /N/ and /L/, unless it follows a voiceless consonant, in which case it surfaces as [т].  An example includes the accusative and plural suffixes:

>>>>

One of the morphotactic challenges met in defining a finite-state transducer for Kyrgyz is that many finite verb forms have ``irregular'' negative forms.  While the paradigms are completely regular, the negative morphotactics are not regularly derived from the affirmative forms.  There are also several different ``regular'' patterns of alternation.  Listed in table~\ref{irregnegs} are a couple of examples of a few finite verb paradigms with their negative forms.

@@

However, instead of doing complicated splitting of the case continuation lexica following the third person possessive suffix to sometimes insert /н/, we decided to proceed under the premise that /н/ was instead underlying in this suffix (and a couple of others!) and got deleted in the nominative, accusative, and genitive.  To accomplish this, a \{n\} archiphoneme was added to the 3rd person possessive suffix, creating an underlying form of \{S\}\{I\}\{n\}.  Phonological rules were implemented in twolc which deleted \{n\} when followed by nothing (for nominative), another set of rules that deleted the accusative \{N\}\{I\} after \{n\} and a morpheme boundary, and a rule that deleted \{n\} when followed by a morpheme boundary and the genitive \{N\}\{I\}н.\footnote{Because a general possessive suffix that can follow personal possessive suffixes can also behave like the genitive in this respect, the rule is actually more general.}

@@

\subsection{Desonorisation}
Desonorisation \citep{washington10} happens to /N/ when it follows any consonant, and to /L/ when it follows a consonant of equal or lower sonority (i.e., /л/, nasals (/м/, /н/, /ң/), and obstruents, but not /й/, /р/, or vowels).  The resulting surface consonant is [д] for both /N/ and /L/, unless it follows a voiceless consonant, in which case it surfaces as [т].  An example includes the accusative and plural suffixes: % FIXME: This should point to the table ?
